From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-08-05 WOLFHPC15/article/component.tex
5.3, discussion modularite solution

Figure~\ref{fig:mscac:assembly} gives an overview of the back-end component
assembly that is executed. As discusses in Section~\ref{sect:mscac},
there are different types of components. Components~\texttt{Driver},
\texttt{DriverApp}, and \texttt{Time} are always instanciated in the final component assembly; the input MSL file is not used by them except the component \texttt{Time} which uses the terminal \texttt{iteration} of the grammar. There is also a
single Component~\texttt{DDS} (the current version handles a single mesh
type). All these components are provided with the
compiler; dumping this part of the assembly is straightforward.

While Component~\texttt{DDS} is managing the structure of the mesh,
each data of the simulation is handled by a component of type
\texttt{Data}. Therefore, the compiler generates in the assembly as
many instances of such component as needed, from the \texttt{data}
section of an MSL program.

The last part of the assembly is the computation part.
Each computation kernel is embedded in a different kernel component which has to be written by the user and provided
to the compiler.
We denote this kind of components K.
A kernel component encapsulates a computations. To
access data, a kernel component is connected, by the compiler, to the adequate
\texttt{Data} components such that their ports name respect the names
given in the \texttt{data} section of MSL. We denote the fact that they may have
several use port by adding a star on the use port, as illustrated in Figure~\ref{fig:k}.
\begin{figure}
\begin{center}
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (k) at (0,0) {$K$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[right=0.2 of u] (star) {*};

@@

%% In the computation part of the simulation a final component is needed,
%% the computation component, denoted K. This type of component
%% represents a single computation of the simulation, i.e. the
%% computation of a single output data (written) using a set of input
%% data (read). Using a \emph{use-multiple}, however, a list of component
%% is used and the explicit identification of each component to use is
%% lost. If for SEQ or PAR, the identification of components is not
%% usefull, it is usefull to manipulate data in a computation. For this
%% reason a computation component has as much \emph{use} ports as input
%% and output data to manipulate. We denote this in the bellow
%% representation as a star on the use port.

The rest of the computation part of the assembly contains the static parallel schedule
computed by MSC, \ie $\Gamma_{tsp}$. This is achieved through the use of
three specific components which manage the $P$, $S$, and $sync$
operations of $\Gamma_{tsp}$ as explained hereafter.

>>>>

\begin{figure}[t]
\begin{center}
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[provide] (Dp) at (-1,0) {};
   \node (Ds) at (-1.5,0) {start};
   \node[component] (D) at (0,0) {$Driver$};
   \node[use,right=1.5cm of D] (Du1) {};
   \node[use,below=0.8cm of D] (Du2) {};
   \node[use,right=1.7cm of Du2] (Du3) {$m$};

   \node[provide,below=0.15 of Du2] (Tp) {};
   \node[component,below=0.8cm of Tp] (T) {$Time$};
   \node[use,below=0.8cm of T] (Tu) {};
   \node[left=0.8cm of T] (tt) {$T$};

   \node[provide,below=0.15 of Tu] (Sp) {};
   \node[component,below=0.8cm of Sp] (S) {$Scheduling$};
   \node[use,right=1.5cm of S] (Su) {};
   \node[use,right=1.5cm of Tu] (Su2) {};
   \node[left=1.5cm of S] (gamma) {$\Gamma_{tsp}$};

   \node[provide,right=0.15 of Su] (Cp) {};
   \node[component,right=1.5cm of Cp] (C) {$Kernels$};
   \node[use,above=0.8cm of C] (Cu) {};
   \node[right=0.2cm of Cu] (star) {$*$};

   \node[provide,right=0.2 of Du3] (Datap1) {};
   \node[provide,above=0.15 of Cu] (Datap2) {};
   \node[provide,right=0.15 of Su2] (Datap3) {};
   \node[component,above=0.8cm of Datap2] (Data) {$Data$};
   \node[use,above=0.8cm of Data] (Datau) {};
   \node[right=1cm of Data] (delta) {$\Delta,\mathcal{D}$};

   \node[provide,right=0.15 of Du1] (DDSp1) {};
   \node[provide,above=0.15 of Datau] (DDSp2) {};
   \node[component,above=0.8cm of DDSp2] (DDS) {$DDS$};
   \node[right=1.2cm of DDS] (m) {$\mathcal{M},\mathcal{E}$};

  \path[-]
    (Dp)     edge node {} (D)
    (D)      edge node {} (Du1)
             edge node {} (Du2)
             edge node {} (Du3)
    (DDSp1)  edge node {} (DDS)
    (T)      edge node {} (Tu)
    (Tp)     edge node {} (T)
    (Sp)     edge node {} (S)
    (S)      edge node {} (Su)
             edge node {} (Su2)
    (Cp)     edge node {} (C)
    (C)      edge node {} (Cu)
    (Datap1) edge node {} (Data)
    (Datap2) edge node {} (Data)
    (Datap3) edge node {} (Data)
    (Data)   edge node {} (Datau)
    (DDSp2)  edge node {} (DDS);
\end{tikzpicture}
\vspace*{.5em}
\caption{MSCAC Component Assembly Back-End Overview.}
\label{fig:mscac:assembly}
\end{center}
\end{figure}

As described in Section~\ref{sect:mscac}, the compiler generates five independent parts in a component assembly in addition to a static part.
This is represented on Figure~\ref{fig:mscac:assembly} where each rounded box represents one or more component instances.
\texttt{Driver} is static and drives the whole application while \texttt{Kernels} contains code directly written by the user.
The four other parts are generated based on the content of sections of the MSL program: \texttt{Scheduling} from $\Gamma$, \emph{DDS} from $\mathcal{M}$ and $\mathcal{E}$, \emph{Data} from $\Delta$ and $\mathcal{D}$, and \emph{Time} from $T$.

\texttt{DDS} manages the structure of the mesh, a single instance is generated as MSL only supports a single mesh for now.
The back-end presented in this paper uses the distributed data structure proposed in SkelGIS library~\cite{CPE:CPE3494}.
\texttt{DDS} is used by \texttt{Data} containing one component instance for each data element in the MSL program.

\texttt{Time} is parametrized by the the \texttt{iteration} value from MSL and calls the \texttt{Scheduling} part as many times as required.
\texttt{Scheduling} contains the static scheduling computed by MSC, \ie $\Gamma_{tsp}$.
The series-parallel tree is encoded using three specific components which manage the $P$, $S$, and $sync$ operations of $\Gamma_{tsp}$ as explained hereafter.
At the leafs of the tree, the components instances make call either to \texttt{Data} for synchronization of to \texttt{Kernels} for computation.

\texttt{Kernels} contains code written by the user to implement the actual computation.
As presented on Figure~\ref{fig:k}, these components provide a single port used to launch the computation and expose one use port for each data element manipulated in the computation.
We denote the fact that they may have several use port by adding a star on the use port.
Each of these use ports is connected to the matching component instance in \texttt{Data}.
We denote this kind of component K.

As a result of this approach, each part of the generated code is rather independent.
Changing the main loop to use a convergence criteria rather than a fixed number of iterations would only require changing the \texttt{Time} component.
Similarly, changing the approach used for scheduling would only impact the \texttt{Scheduling} components.
The technology used for data parallelism can also be changed by replacing the \texttt{DDS} and \texttt{Data} components; as a matter of fact work is in progress to propose an alternative based on Global Arrays~\cite{Nieplocha:2006:AAP:1125980.1125985} instead of SkelGIS.
In this case however, the \texttt{Kernels} components written by the user have to be adapted to the new interface, but this is the only impact.

\begin{figure}
\begin{center}
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (k) at (0,0) {$K$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[right=0.2 of u] (star) {*};

@@

### 2015-08-05 WOLFHPC15/article/mscac.tex
5.3, discussion modularite solution

Thanks to the software components based approach, the generated code is made of independent parts:
A first set of components encodes the static scheduling provided by MSC.
The other parts loosely match the elements of the multi-stencil program definition, these include the components implementing: \textbf{a)} the kernels ($\Gamma$), \textbf{b)} the data structures ($\mathcal{M}$ and $\mathcal{E}$), \textbf{c)} their parallelization ($\Delta$ and $\mathcal{D}$), and \textbf{d)} the time loop ($T$).
This enables to change the technology used for each part while limiting the impact on the other parts as further discussed in Section~\ref{sect:component}.

>>>>

Thanks to the software components based approach, the generated code is made of five independent parts that loosely match the elements of the multi-stencil program definition.
These are the components implementing: \textbf{a)} the scheduling ($\Gamma$), \textbf{b)} the data structures ($\mathcal{M}$ and $\mathcal{E}$), \textbf{c)} their parallelization ($\Delta$ and $\mathcal{D}$), \textbf{d)} the time loop ($T$), and \textbf{e)} the kernels directly provided by the user.
This enables to change the technology used for each part while limiting the impact on the other parts as further discussed in Section~\ref{sect:component}.
